\documentclass[11pt, a4paper]{article}
\usepackage{fontspec}
\usepackage[round]{natbib}
\usepackage[margin=0.5in]{geometry}
%usepackage[top=0.5in, bottom=0.5in, left=0.5in, right=0.5in]{geometry}
%\usepackage{acl}
\usepackage{array}
\usepackage{booktabs}
\usepackage[normalem]{ulem}
\usepackage[colorlinks=true, urlcolor=blue, linkcolor=blue]{hyperref}
\usepackage{float}

\setmainfont{Arial}

\title{Detecting Machine Generated Texts in isiZulu \\[1ex] \large Evaluating pre-trained models and diverse datasets}
\author{Ruan Esterhuizen (u23532387), Ruben Hannes Gadd (u23633353), Christiaan Strydom (u22498878) \\[1ex] \large Group 19}
\date{}

\begin{document}
    \maketitle
    \section{Introduction}
    With the dawn of the 20th decade of the 21st millennium, the introduction of AI models such as ChatGPT has caused rippling 
    effects across the world both positive and negative. Such negative effects include the abuse of these models to 
    generate texts that are becoming increasingly difficult to differentiate between text written by man. In response 
    to these issues, NLP techniques are utilized with the purpose of identifying between MGT (Machine generated Text) and 
    HGT (Human Generated Text). In this report proposal we detail an approach to creating a Zulu-based MGT detection tool. 
    We chose Zulu/isiZulu since it is the most spoken African language in South Africa and shares linguistic roots with other African
    languages that fall under the Nguni linguistic tree which include isiXhosa, isiSwati and isiNdebele. Additionally, since isiZulu is
    not as globally spoken outside of Africa, not many models that focus on such languages exist. Note throughout this proposal the 
    use of MGT (Machine generated Text) and HGT (Human Generated Text) will be used.

    \section{Background}
    For our proposal we consulted previous work performed that also aimed to create a model for MGT detection. The paper "Who Wrote This? 
    Identifying Machine vs Human-Generated Text in Hausa"\citep{hausa} had the aim of developing a MGT detector to identify MGT among HGT written in the 
    Afro-asiatic low resource language of Hausa, which was achieved by fine-tuning four pre-trained Afri-centric models on a dataset constituting 
    a mixture of News articles in Hausa and MGT that was generated via Gemini. Of the four models the AfroXLMR model was reported to perform the best. A 
    limitation of the study that is hoped to be addressed in this proposal is that the HGT used only consisted of News articles with no consideration of 
    the use of other sources of text. Additionally, the study also only relied on MGT generated by one source without considering other alternative sources. 
    Another consulted paper is "Arabic machine-generated text detection: Stylometric analysis and cross-model evaluation"\citep{alshaibani2026arabic} which investigated the development of 
    a MGT detector for the high resource language of Arabic. In the paper the model was trained on MGT based on formal papers and social media that was 
    generated via various LLMs. When evaluating the model's performance, the researchers found that the model performed worse on MGT based off social media. This 
    was due to how MGT and HGT written in a casual context start to resemble each other. The paper further suggested that to overcome this limitation, the MGT used 
    for training must also be further generalised by changing the prompts used to generate each MGT.
    \section{Methodology}

\subsection{Research Questions}

\begin{itemize}
    \item How do different multilingual models compare in MGT detection in isiZulu?
    \item What linguistic and stylistic features can be seen in isiZulu MGTs?
    \item Does diversity in training data enhance detector generalizability?
\end{itemize}

\subsection{Datasets}

\begin{table}[h]
\centering
\begin{tabular}{l l l >{\raggedright\arraybackslash}p{5cm} l}
\toprule
\textbf{Name} & \textbf{Domain} & \textbf{Size (items)} & \textbf{Description} & \textbf{Repository} \\
\midrule
Vuk'uzenzele 
    & Magazine articles 
    & 185 
    & Articles from SA government magazine 
    & \href{https://github.com/dsfsi/vukuzenzele-nlp}{\uline{link}} \\
IsiZulu News 
    & News articles 
    & 752 
    & Scraped publicly available news articles
    & \href{https://github.com/dsfsi/za-isizulu-siswati-news-2022}{\uline{link}} \\
AfriHate* 
    & Social media 
    & $\sim$4000 
    & Scraped public Twitter posts 
    & \href{https://github.com/AfriHate/AfriHate}{\uline{link}} \\
\bottomrule
\end{tabular}
\caption{Datasets}
\end{table}

\textbf{Preprocessing}: All data subsets will have the same size to prevent biases in training. Only isiZulu entries will be used for multilingual datasets.

All datasets will be combined into a singular dataset containing the text and the corresponding label (MGT/HGT).

All datasets are licensed for their use for this project.

* This dataset contains sensitive content; however, only benign entries will be used.



\subsection{Approach and Methods}

\subsubsection{Models}
\begin{table}[h]
\centering
\begin{tabular}{l l l l}
\toprule
\textbf{Name} & \textbf{Languages} & \textbf{Size (parameters)} & \textbf{Repository} \\
\midrule
XLM-RoBERTa-base
    & 100
    & ~125M
    & \href{https://huggingface.co/docs/transformers/en/model_doc/xlm-roberta}{\uline{link}} \\
AfroXLMR-large
    & 17
    & 559.9M
    & \href{https://huggingface.co/Davlan/afro-xlmr-large}{\uline{link}} \\
Xhosa\_ZuluLlama3
    & 2 
    & 8B 
    & \href{https://huggingface.co/Jacaranda/Xhosa_ZuluLlama3_v1}{\uline{link}} \\
\bottomrule
\end{tabular}
\caption{Models}
\end{table}

\subsubsection{Approach}
A diverse dataset, compromising of multiple domains of HGT (e.g. news and social media) will be used. Moreover, MGTs shall be generated using a variety of LLMs (e.g. ChatGPT, Gemini). Dataset diversity has been noted as a limitation in previous work.


Models trained on varying numbers of languages (see Table 2) will be used to examine the impact of linguistic diversity.


This approach is expected to improve the accuracy and generalizability of MGT detection relative to previous studies.


\subsection{Work Plan}
\subsubsection{Timeline}

\begin{table}[H]
\centering
\begin{tabular}{l l >{\raggedright\arraybackslash}p{8.5cm} l}
\toprule
\textbf{Phase} & \textbf{Task} & \textbf{Description} & \textbf{*Deadline}\\
\midrule
1
    & Dataset Generation
    & Each member collects LLM data for each domain
    & 4 May \\
2 
    & LLM Fine Tuning
    & Each member fine-tunes their assigned model
    & 18 May \\
3 
    & Evaluation
    & Models are tested and results are reported upon
    & 25 May  \\
\bottomrule
\end{tabular}
\caption{Proposed Project Timeline}
\end{table}

\subsubsection{Feasibility \& Risk Assessment}
The work plan will be adhered to through internal deadlines*. Major risks associated include scope creep and potential delays, which will be managed via regular communication.

* Deadlines are subject to change
    
    \section{Evaluation}
    To evaluate the performance of our proposed models in detecting machine-generated text (MGT) and assessing textual similarity, we utilize standard classification metrics, namely Accuracy, Precision, Recall, and F1-score, and reference-based similarity metrics, namely BLEU \citep{papineni2002bleu}, METEOR \citep{banerjee2005meteor}, ROGUE-L \citep{lin2004rouge}, and BERTScore \citep{zhang2019bertscore}.

For classification performance, Accuracy measures the overall correct classifications, while Precision measures the proportion of True Positive(TP) predictions to the total predicted positives. Recall measures the model's ability to detect True Positive predictions from all True Positive predictions available. The F1-score provides a harmonic mean of Precision and Recall, providing a balance between the two metrics.

The three reference-based similarity metrics used for syntactic similarity are BLEU which measures precision based on exact word/phrase overlap, METEOR which measures a balance of recall and precision allowing synonyms and stemming, and ROUGE-L which measures recall by finding the longest common subsequences of words between the generated and human text. The metric measure semantic similarity is BERTScore which measures similarity using Artificial Intelligence(AI) model embeddings rather than exact word matches.

Furthermore, to evaluate the generalisation capabilities of the MGT detection models, we will test them on an unseen dataset drawn from a different domain. This dataset is sourced from Zenodo \citep{zenodo5035171}, enabling evaluation under cross-domain and cross-model conditions.

    \section{Expected Outcome}
The expected outcome of this study is to produce a mode effectively detecting Zulu machine-generated text (MGT), which contributes to low-resource language detection research. For future work, the model’s generalisation to languages with similar linguistic structures can be evaluated.

We anticipate several challenges, including the lack of native Zulu expertise, limited quantities of Zulu text data due to its low-resource nature, and reduced generalisation across different topic domains and MGT sources.

\bibliographystyle{plainnat}
%\bibliographystyle{acl}
\bibliography{references}
    \end{document}
